---
title: "Assignment 1 Submission"
author: "Prajen Maharjan"
date: "submission date"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statement of Contribution

Include this if pairing.

State what each team member has contributed to the assignment before answering the questions. Each student must contribute towards ALL questions. Your assignment will not be marked without this statement.

```{r eval=FALSE, include=FALSE}
## There are a few options in every chunk, click the setting button (greyed) near the triangle:
# 1. to use default as in the above (show code and output, ie ECHO=TRUE) or 
# 2. to show output only (ie ECHO=FALSE) or
# 3. show nothing (run code) ie include=FALSE
# 4. show nothing (dont run the code), like this chunk, ie eval=FALSE, include=FALSE
## You can knit this template into Word and then update it into pdf etc..
## You can create your own way for submission, this is just an example if you are going to use Rmd.
```


# Question 1 Normal distribution. (8 marks) 


(a) (1 mark) Find the probability that the user spends more than 15 minutes per month at the site. 

    **Answer:**

0.9937

```{r}
# Answer code here
1 - pnorm (15, 25, 4)
```

(b) (2 marks) Find the probability that the user spends between 20 and 35 minutes per month at the site.

    **Answer:**

0.8881
   
```{r}
probability <- pnorm(35, 25, 4) - pnorm(20, 25, 4)
probability
0.8881406
```

(c) (2 marks) What is the amount of time per month a user spends on Facebook, if only 1\% of users spend this time or longer on Facebook? 

    **Answer:**
    
34.3053

```{r}
time_spent <- qnorm(1- 0.01, 25, 4)
time_spent
34.30539
```

(d) (3 marks) Between what values do the time spent of the middle 90\% distribution of Facebook users fall? 

    **Answer:**
    
The values lies between 18.4205 and 31.5794

```{r}
lower_bound <- qnorm(0.05, 25, 4)
upper_bound <- qnorm(0.95, 25, 4)
lower_bound
18.42059
upper_bound
31.57941
```


# Question 2 Blood fat concentration (11 marks)

(a) (6 marks) Conduct a two-independent sample $t$-test using R to determine whether the concentration of plasma cholesterol is significantly different between patients with no evidence of heart disease and those with narrowing of the arteries.

    **Answer:**

-Step 1. State the hypotheses. 
H0 : µ1 = µ2 against H1 : µ1 ≠ µ2

where µ1 and µ2 are the population means of patients with heart disease and other with no such disease respectively
-Step 2 the test statistic is t = -4.4323 and
-Step 3 the sampling distribution is df=69.113. 
-Step 4 p-value = 3.42e-05 i.e 0.0000342 > 0.01.
-Step 5 since the p-value < 0.01 (1% significance level), we reject the null hypothesis (H0)
-Step 6 We conclude that there is a statistically significant difference in cholesterol levels between the two groups.

```{r}
mean_no_disease <- 195.2745
mean_disease <- 216.1906
var_no_disease <- 1303.9231
var_disease <- 1850.2488
n_no_disease <- 51
n_disease <- 320

group1 <- rnorm(n_no_disease, mean_no_disease, sqrt(var_no_disease))
group2 <- rnorm(n_disease, mean_disease, sqrt(var_disease))
t_test_result <- t.test(group1, group2, var.equal = FALSE, conf.level = 0.99)

t_test_result
```


(b) (3 marks) Determine a 99\% confidence interval for the mean difference in concentration of plasma cholesterol between the two groups of patients.

    **Answer:**

    *Answer text here*

(c) (2 marks) Explain the correspondence between the confidence interval in (b) and a test of the hypotheses you listed in question (a).

    **Answer:**

    *Answer text here*

# Question 3 Regression (31 marks) 

(a) (2 marks) Fit a simple linear model $M_1$ to these data. Present the appropriate scatterplot and plot the fitted line onto the scatterplot. Comment about the output in a few concise sentences.

    **Answer:**

    *Answer text here*

(b) (5 marks) Provide the model summary and diagnostics checking plots for model $M_1.$ Does the straight line regression model $M_1$ seem to fit the data well? Comment about the output in a few concise sentences. 

    **Answer:**

    *Answer text here*

(c) (5 marks) Do you think there are outliers or influential points in the data? What influence do these points have on the model fit? Use leverage and Cook's distance for this investigation. 

    **Answer:**

    *Answer text here*

(d) (4 marks) Fit a regression model to the transformed $M_2$ model. Present the appropriate scatterplot and plot the fitted line onto the scatterplot. Does the transformed line regression model $M_2$ seem to fit the data well? Comment about the output in a few concise sentences.

    **Answer:**

    *Answer text here*

(e) (5 marks) Provide the model summary and diagnostics checking plots for model $M_2.$ Does the straight line regression model $M_2$ seem to fit the data well? Comment about the output in a few concise sentences.

    **Answer:**

    *Answer text here*

(f) (4 marks) Perform a hypothesis testing for a positive slope at a significance level of 5\% based on model $M_2.$

    **Answer:**

    *Answer text here*

(g) (6 marks) Compare a 95\% confidence interval of the mean response and a 95\% prediction interval for a new value when Tonnage = 10,000 using the untransformed model $M_1$ and transformed model $M_2$ respectively. Provide two scatterplots that consist the fitted model, the confidence and prediction intervals for each of $M_1$ and $M_2$ respectively. Comment about the output in a few concise sentences.

    **Answer:**

    *Answer text here*
