---
title: "Assignment 1 Submission"
author: "Prajen Maharjan  (24841288)"
date: "11 April, 2025"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statement of Contribution

Include this if pairing.

State what each team member has contributed to the assignment before answering the questions. Each student must contribute towards ALL questions. Your assignment will not be marked without this statement.

```{r eval=FALSE, include=FALSE}
## There are a few options in every chunk, click the setting button (greyed) near the triangle:
# 1. to use default as in the above (show code and output, ie ECHO=TRUE) or 
# 2. to show output only (ie ECHO=FALSE) or
# 3. show nothing (run code) ie include=FALSE
# 4. show nothing (dont run the code), like this chunk, ie eval=FALSE, include=FALSE
## You can knit this template into Word and then update it into pdf etc..
## You can create your own way for submission, this is just an example if you are going to use Rmd.
```


# Question 1 Normal distribution. (8 marks) 


(a) (1 mark) Find the probability that the user spends more than 15 minutes per month at the site. 

    **Answer:**

```{r}
1 - pnorm (15, 25, 4)
```

    The probability that the user spends more than 15 minutes per month at the site is 99.3790%


(b) (2 marks) Find the probability that the user spends between 20 and 35 minutes per month at the site.

    **Answer:**
   
```{r}
probability <- pnorm(35, 25, 4) - pnorm(20, 25, 4)
probability
```

      The probability that the user spends between 20 and 35 minutes per month at the site is 88.8140%

(c) (2 marks) What is the amount of time per month a user spends on Facebook, if only 1\% of users spend this time or longer on Facebook? 

    **Answer:**
    
```{r}
qnorm(1- 0.01, 25, 4)
```

        The amount of time per month a user spends on Facebook, if only 1\% of users spend this time or longer on Facebook is 34.3053


(d) (3 marks) Between what values do the time spent of the middle 90\% distribution of Facebook users fall? 

    **Answer:**

```{r}
lower_bound <- qnorm(0.05, 25, 4)
upper_bound <- qnorm(0.95, 25, 4)
lower_bound
upper_bound
```

      The values lies between 18.4205 and 31.5794

# Question 2 Blood fat concentration (11 marks)

(a) (6 marks) Conduct a two-independent sample $t$-test using R to determine whether the concentration of plasma cholesterol is significantly different between patients with no evidence of heart disease and those with narrowing of the arteries.

    **Answer:**

```{r}
  set.seed(123)
  mean_no_disease <- 195.2745
  mean_disease <- 216.1906
  var_no_disease <- 1303.9231
  var_disease <- 1850.2488
  n_no_disease <- 51
  n_disease <- 320
  group1 <- rnorm(n_no_disease, mean_no_disease, sqrt(var_no_disease))
  group2 <- rnorm(n_disease, mean_disease, sqrt(var_disease))
  t_test_result <- t.test(group1, group2, var.equal = FALSE, conf.level = 0.99)
  t_test_result

```

    The six steps of hypothesis testing is carried out as:
    
    Step 1. State the hypotheses. 
    H0 : µ1 = µ2 against H1 : µ1 ≠ µ2
    
    where µ1 and µ2 are the population means of patients with heart disease and other with no such disease respectively
    Step 2 the test statistic is t = -4.0456 and
    Step 3 the sampling distribution is df=78.352. 
    Step 4 p-value = 0.0001176 < 0.01.
    Step 5 since the p-value < 0.01 (1% significance level), we reject the null hypothesis (H0)
    Step 6 We conclude that there is a statistically significant difference in cholesterol levels between the two groups.


(b) (3 marks) Determine a 99\% confidence interval for the mean difference in concentration of plasma cholesterol between the two groups of patients.

    **Answer:**

```{r}
t_test_result$conf.int
```

    From the test output, the 99% confidence interval for the difference in means is:
    [−34.82, −7.36]
    
    We are 99% confident that the true difference in mean cholesterol levels lies between -34.82 mg/dL and -7.36 mg/dL.
    Since the entire interval is negative, it indicates that patients with narrowing of the arteries tend to have higher plasma cholesterol levels than those with no disease.

(c) (2 marks) Explain the correspondence between the confidence interval in (b) and a test of the hypotheses you listed in question (a).

    **Answer:**

    The mean difference which we calculated using R at 99% confidence interval for the mean difference does not include 0, which shows significant difference. This supports the hypothesis test carried out on part a, where we rejected the null hypothesis.

    Thus, both part b and c lead to the same conclusion: There is a statistically significant difference in plasma cholesterol levels between the two patient groups.


# Question 3 Regression (31 marks) 

(a) (2 marks) Fit a simple linear model $M_1$ to these data. Present the appropriate scatterplot and plot the fitted line onto the scatterplot. Comment about the output in a few concise sentences.

    **Answer:**
    
```{r}
data <- read.table("glakes.txt", header = TRUE)
model1 <- lm(Time ~ Tonnage, data = data)
plot(data$Tonnage, data$Time, main = "Tonnage vs Time", xlab = "Tonnage", ylab = "Time")
abline(model1, col = "blue", lwd = 2)
```

    Conclusion: 
    
    The above scatterplot shows a positive relationship between tonnage and time. It shows that ships with higher tonnage tend to spend more time in port.

(b) (5 marks) Provide the model summary and diagnostics checking plots for model $M_1.$ Does the straight line regression model $M_1$ seem to fit the data well? Comment about the output in a few concise sentences. 

    **Answer:**
    
```{r}
data <- read.table("glakes.txt", header = TRUE)
fit_glakes <- lm(Time ~ Tonnage, data = data)
summary(fit_glakes)
par(mfrow = c(2, 2))
plot(fit_glakes)
```
    
    The model summary is:

    Call:
    lm(formula = Time ~ Tonnage, data = glakes)
    
    Residuals:
        Min      1Q  Median      3Q     Max 
    -23.882  -6.397  -1.261   5.931  21.850 
    
    Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
    (Intercept) 12.344707   2.642633   4.671 6.32e-05 ***
    Tonnage      0.006518   0.000531  12.275 5.22e-13 ***
    ---
    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
    
    Residual standard error: 10.7 on 29 degrees of freedom
    Multiple R-squared:  0.8386,	Adjusted R-squared:  0.833 
    F-statistic: 150.7 on 1 and 29 DF,  p-value: 5.218e-13

    Conclusion: 
    
    The linear regression model M1 shows a statistically significant relationship between Tonnage and Time (p = 5.22e-13). The coefficient for Tonnage (0.0065) indicates that for each additional ton of cargo, loading time increases by approximately 0.0065 hours.
    
(c) (5 marks) Do you think there are outliers or influential points in the data? What influence do these points have on the model fit? Use leverage and Cook's distance for this investigation. 

    **Answer:**

```{r}
glakes <- read.table("glakes.txt", header = TRUE)
fit_glakes <- lm(Time ~ Tonnage, data = data)
par(mfrow = c(2, 2)) 
plot(fit_glakes)
glakes_infl <- influence.measures(fit_glakes)
leverage_inv <- hatvalues(fit_glakes)
cooks_dist_inv <- cooks.distance(fit_glakes)
leverage_inv
cooks_dist_inv
plot(fit_glakes, which = 5, main = "Residuals vs Leverage")
plot(fit_glakes, which = 4, main = "Cook’s Distance by Case")
```

    Firstly, Analysing Residuals vs Leverage plot, we get: 
    
    For  n=31, hii>4/31≈0.13
    Case 31 (leverage = 0.416) exceeds this threshold (far right on the plot). Case 3 (top-center) has high leverage but moderate influence.
    
    Secondly, Analysing Cook's distance, we get: 
    
    Case 31 falls outside the outermost contour (Cook’s D = 1.20), confirming high influence.
    
    In Conclusion:
    
    1. Outliers: Cases 3 (std_residual = 2.15) and 6 (std_residual = 1.98).
    2. High Leverage: Cases 3 (leverage = 0.222) and 31 (leverage = 0.416).
    3. Influential Points: Case 31 (Cook’s D = 1.20).
    
    The influence inflates R² in the original model.

(d) (4 marks) Fit a regression model to the transformed $M_2$ model. Present the appropriate scatterplot and plot the fitted line onto the scatterplot. Does the transformed line regression model $M_2$ seem to fit the data well? Comment about the output in a few concise sentences.

    **Answer:**
    
```{r}
glakes <- read.table("glakes.txt", header = TRUE)
fit_m2 <- lm(log(Time) ~ sqrt(Tonnage), data = glakes)
plot(log(Time) ~ sqrt(Tonnage), data = glakes, main = "Transformed Model M2: log(Time) vs sqrt(Tonnage)", xlab = "Square Root of Tonnage", ylab = "log(Time)")
abline(fit_m2, col = "red", lwd = 2)
summary(fit_m2)
```
    In conclusion:
    
      The transformed model M2  shows a strong, statistically significant relationship (p = 4.1e-12 for the slope). The model explains 81.4% of the variance in log(Time), indicating a good fit. For every 1-unit increase in sqrt(Tonnage), log(Time) increases by ~0.0215.
    

(e) (5 marks) Provide the model summary and diagnostics checking plots for model $M_2.$ Does the straight line regression model $M_2$ seem to fit the data well? Comment about the output in a few concise sentences.

    **Answer:**
    
```{r}
glakes <- read.table("glakes.txt", header = TRUE)
m2 <- lm(log(Time) ~ sqrt(Tonnage), data = glakes)
summary(m2)
```
    
    The model summary is:

    Call:
    lm(formula = log(Time) ~ sqrt(Tonnage), data = glakes)
    
    Residuals:
        Min      1Q  Median      3Q     Max 
    -0.6408 -0.2522 -0.0357  0.2457  0.5814 
    
    Coefficients:
                  Estimate Std. Error t value Pr(>|t|)    
    (Intercept)   2.210424   0.111580   19.81  < 2e-16 ***
    sqrt(Tonnage) 0.021514   0.001909   11.27  4.1e-12 ***
    ---
    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
    
    Residual standard error: 0.3048 on 29 degrees of freedom
    Multiple R-squared:  0.8141,	Adjusted R-squared:  0.8077 
    F-statistic:   127 on 1 and 29 DF,  p-value: 4.098e-12
    
    In conclusion: 
    
    The transformed model M2 fits the data very well i.e  (R² = 0.814) with significant coefficients (p < 0.001). Residuals are linear, normally distributed. M2 also does not show any severe leverage or residuals.


(f) (4 marks) Perform a hypothesis testing for a positive slope at a significance level of 5\% based on model $M_2.$

    **Answer:**

    The six steps of hypothesis testing is carried out as:
    
    Step 1. State the hypotheses.
    H₀: β₁ ≤ 0 (The slope of sqrt(Tonnage) is zero or negative).
    H₁: β₁ > 0 (The slope is positive, implying sqrt(Tonnage) increases log(Time)).
    Step 2. The test statistic is: t = 11.27, derived from the estimated slope (β₁ = 0.0215) and its standard error (SE = 0.0019).
    Step 3. The sampling distribution is: t-distribution with df = 29 (degrees of freedom = n − 2 = 31 − 2).
    Step 4. p-value = 4.1 × 10⁻¹² (from the model summary), which is < 0.05.
    Step 5. Decision: Since the p-value < 0.05 (5% significance level), we reject H₀.
    Step 6. Conclusion:
    There is statistically significant evidence (at α = 0.05) to conclude that the slope is positive. This confirms that as ship tonnage increases, loading/unloading time (log(Time)) significantly increases.


(g) (6 marks) Compare a 95\% confidence interval of the mean response and a 95\% prediction interval for a new value when Tonnage = 10,000 using the untransformed model $M_1$ and transformed model $M_2$ respectively. Provide two scatterplots that consist the fitted model, the confidence and prediction intervals for each of $M_1$ and $M_2$ respectively. Comment about the output in a few concise sentences.

    **Answer:**
    For M1 

```{r}
glakes <- read.table("glakes.txt", header = TRUE)
 m1 <- lm(Time ~ Tonnage, data = glakes)
 newdata <- data.frame(Tonnage = 10000)
 m1_pred <- predict(m1, newdata, interval = "confidence", level = 0.95)
 m1_pred_pred <- predict(m1, newdata, interval = "prediction", level = 0.95)
 m1_pred
 m1_pred_pred
plot(glakes$Tonnage, glakes$Time, main = "Model M1: Time ~ Tonnage", xlab = "Tonnage", ylab = "Time", pch = 19, col = "blue")
abline(m1, col = "red")
tonnage_seq <- seq(min(glakes$Tonnage), max(glakes$Tonnage), length.out = 100)
m1_conf <- predict(m1, newdata = data.frame(Tonnage = tonnage_seq), interval = "confidence")
m1_pred_band <- predict(m1, newdata = data.frame(Tonnage = tonnage_seq), interval = "prediction")
lines(tonnage_seq, m1_conf[, "lwr"], col = "red", lty = 2)
lines(tonnage_seq, m1_conf[, "upr"], col = "red", lty = 2)
lines(tonnage_seq, m1_pred_band[, "lwr"], col = "green", lty = 2)
lines(tonnage_seq, m1_pred_band[, "upr"], col = "green", lty = 2)
legend("topleft", legend = c("Data", "Fitted", "CI (mean)", "PI (new)"), col = c("blue", "red", "red", "green"), lty = c(NA, 1, 2, 2), pch = c(19, NA, NA, NA))

```

    For M2
  
```{r}
 glakes <- read.table("glakes.txt", header = TRUE)
 m2 <- lm(log(Time) ~ I(Tonnage^0.5), data = glakes)
 newdata <- data.frame(Tonnage = 10000)
 m2_pred_log <- predict(m2, newdata, interval = "confidence", level = 0.95)
 m2_pred <- exp(m2_pred_log)
 m2_pred
 m2_pred_pred_log <- predict(m2, newdata, interval = "prediction", level = 0.95)
 m2_pred_pred <- exp(m2_pred_pred_log)
 m2_pred_pred
 plot(glakes$Tonnage, log(glakes$Time), main = "Model M2: log(Time) ~ sqrt(Tonnage)",  xlab = "Tonnage", ylab = "log(Time)", pch = 19, col = "blue")
 tonnage_seq <- seq(min(glakes$Tonnage), max(glakes$Tonnage), length.out = 100)
 pred_log <- predict(m2, newdata = data.frame(Tonnage = tonnage_seq))
 lines(tonnage_seq, pred_log, col = "red", lwd = 2) 
 conf_log <- predict(m2, newdata = data.frame(Tonnage = tonnage_seq), interval = "confidence")
 pred_band_log <- predict(m2, newdata = data.frame(Tonnage = tonnage_seq), interval = "prediction")
 lines(tonnage_seq, conf_log[, "lwr"], col = "red", lty = 2)
 lines(tonnage_seq, conf_log[, "upr"], col = "red", lty = 2)
 lines(tonnage_seq, pred_band_log[, "lwr"], col = "green", lty = 2) 
 lines(tonnage_seq, pred_band_log[, "upr"], col = "green", lty = 2)
 legend("topleft",  legend = c("Data", "Fitted", "CI (mean)", "PI (new)"),  col = c("blue", "red", "red", "green"), lty = c(NA, 1, 2, 2), pch = c(19, NA, NA, NA))

```

    In conclusion:
    1. M2's intervals (CI: 62.8–97.9, PI: 40.5–151.9) are tighter than M1's (CI: 69.4–85.7, PI: 54.2–100.9), indicating better predictive accuracy with transformations.
    2. M2's curve fits the nonlinear trend better than M1's straight line.
    3. M2 is preferable for port planning due to its robust handling of scale and variance.

  
   
